Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa:Development of a 3D spectral numerical coastal-flow model

### 2019-11-14 literature_review.Rmd
added section of ocean modelling without references. citations will be added soon

---
title: "Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa:Development of a 3D spectral numerical coastal-flow model"
author: 
- RM Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  html_document:
    df_print: paged
    toc: yes
    theme:  simplex
  word_document:
    fig_caption: yes
    reference_docx: chapter_1_template.docx
    toc: yes
  pdf_document:
    toc: no
bibliography: references.bib
---

@@

Ocean models are popular tools for investigating future climate scenarios and has been used mostly in this regard. However, ocean models can also be used in operational oceanography (now-casts and forecasts) in the maritime and shipping industry, experimental oceanography, and can be used to mechanistically interpret ocean observations. Ocean models are similar to atmospheric models but fundamentally differ in some respects. For example, atmospheric models involve air which is a compressible gas while ocean models involve seawater which is nearly in-compressible. These differences have significant impacts for how aspects of volume, temperature and pressure influence one another. In air, the interplay between these factors is largely linear and well understood, while in the ocean the relationship is non-linear and poorly understood. For ocean models the equation state used is $$\rho = f(x)TSP$$
where, $\rho$ represents density and is a function of, $T$ temperature, $S$ salinity, and $P$ pressure. Although the equation state is more complex compared to that of air some aspects are slightly simpler, as seawater is in-compressible the water entering a grid box will be the same as the water exiting the grid box. Another major difference between ocean and atmospheric models is in ocean models salinity must be taken into account while with atmospheric models humidity is an important factor. Ocean and atmospheric models also differ slighlty in terms of vertical structure i.e. layering. The surface of the ocean is where most heating and coolong occurs which cause the formation of a marine boundry or mixed layer. Below the mixed layer the ocean becomes more stratified which is ideal for determining flow. However, complications arise in the horizonal layer which does not continiously layer like the atmosphere. The horizontal layer in ocean models needs to take into account irregular bathymetry, shape and size of continental shelves/basins. Furthermore, run-off from rivers can cause changes in the density of seawater along the coasts, thereby altering their flow regimes. Therefore, ocean models need to resolve both the vertical and horizontal structuring, particulary along the ocean margins, which can be achieved through selecting the correct horizontal and vertical co-ordinate systems. Finally, the ocean interior is largely driven by density gradients compared to the atmosphere which is well mixed and so the equalibrium timescales are much slower relative to the atmosphere. For instance, an ocean model can take as long as 1000 years to start an ocean model from rest.

Ocean models use similar equations to atmospheric models, with some minor differences (might add equation list). Theses equations form the basis of a ocean model, but more input is needed for a model to run successfully. Ocean models also need input on boundary conditions such as basin geometry, bathymetry and atmospheric pressure. The Atmospheric pressure can be solved for dynamically, or at the very least, the effects are approximated and imposed. Initial abiotic conditions are another important input into ocean models. Initial conditions include the mean state of temperature, salinity and velocity fields and can be calculated from climatology data or from a previously run ocean model. Finally, the forcing fields, both dynamic and static, are needed such as shortwave radiation, long-wave radiation, latent heat, evaporation, precipitation and land surface run-off. Examples of dynamic forcing fields include winds and tides.

A model can be defined by the resolution, parametrisation and intergration.

>>>>

---
title: "Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa:Development of a 3D spectral numerical coastal-flow model"
author: 
- RM Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  pdf_document:
    toc: no
  html_document:
    df_print: paged
    toc: yes
    theme:  simplex
  word_document:
    fig_caption: yes
    reference_docx: chapter_1_template.docx
    toc: yes
bibliography: references.bib
---

@@

Ocean models are popular tools for investigating future climate scenarios, but have a range of applications outside the realm of academia. Ocean models can also be used in operational oceanography in the maritime and shipping industry (now-casts and forecasts), experimental oceanography, and can be used to mechanistically interpret ocean observations. Ocean models are similar to atmospheric models but fundamentally differ in some respects. For example, atmospheric models involve air which is a compressible gas while ocean models involve seawater which is nearly in-compressible. These differences have significant impacts for how aspects of volume, temperature and pressure influence one another. In air, the interplay between these factors is largely linear and well understood, while in the ocean the relationship is non-linear and poorly understood. For ocean models the equation state used is $$\rho = f(x)TSP$$
where, $\rho$ represents density and is a function of, $T$ temperature, $S$ salinity, and $P$ pressure. Although the equation state is more complex compared to that of air, some aspects are slightly simpler as seawater is in-compressible and therefore the water entering a grid box will be the same as the water exiting a grid box. Another major difference between ocean and atmospheric models is in ocean models salinity must be taken into account, while with atmospheric models humidity is an important factor. Ocean and atmospheric models also differ slighlty in terms of vertical structure i.e. layering. The surface of the ocean is where most heating and coolong occurs which cause the formation of a marine boundry or mixed layer. Below the mixed layer the ocean becomes more stratified which is ideal for determining flow. However, complications arise in the horizonal layer which does not continiously layer like the atmosphere. The horizontal layer in ocean models needs to take into account irregular bathymetry, shape and size of continental shelves/basins. Furthermore, run-off from rivers can cause changes in the density of seawater along the coasts, thereby altering their flow regimes. Therefore, ocean models need to resolve both the vertical and horizontal structuring, particulary along the ocean margins, which can be achieved through selecting the correct horizontal and vertical coordinate systems. Finally, the ocean interior is largely driven by density gradients compared to the atmosphere which is well mixed and so the equalibrium timescales are much slower relative to the atmosphere. For instance, an ocean model can take as long as 1000 years to start from rest.

Ocean models use similar equations to atmospheric models, with some minor differences (might add equation list). Theses equations form the basis of a ocean model, but more input is needed for a model to run successfully. Ocean models also need information on boundary conditions such as basin geometry, bathymetry and atmospheric pressure. The Atmospheric pressure can be solved for dynamically, or at the very least, the effects are approximated and imposed. Initial abiotic conditions are another important input into ocean models. Initial conditions include the mean state of temperature, salinity and velocity fields and can be calculated from climatology data or from a previously run ocean model. Finally, the forcing fields, both dynamic and static, are needed such as shortwave radiation, long-wave radiation, latent heat, evaporation, precipitation and land surface run-off. Examples of dynamic forcing fields include winds and tides. One of the most important steps in ocean modelling id defining the horizontal and vertical grids to be used. In terms of horizontal grids, there are two main types, namely regular and irregular grids. As the name suggests, a regular grid contains cells of the same size and dimensions i.e the grid cells are evenly spaced. When a regular grid is placed over a spherical earth, the grid cells vary in size and dimension resulting in a curvilinear grid. In addition, the grid lines converge at poles making modelling in those regions almost impossible. A popular solution is to use a tripolar grid laid over the Artic polar region which results in two poles positioned over land, creating more even spacing at the convergence points. The advantage of using regular grids is the decades of research on the subject, relatively straight forward analysis algorithms, and is computationally efficient. Problems arise with regular grids when increasing the resolution over a specific region, as the resolution must be increased along the entire latitude length. This leads to modelling areas not of interest (middle of the ocean) and is a waste of computational resources. As a result, the use of irregular grids in ocean modelling has been growing in popularity in recent years, as the spatial resolution can vary throughout the model.

There are different 'types' of irregular grids available such as using triangles meshed together to form a grid instead of rectangles or squares and is known as 'finite element modelling'. The size of the triangles can be altered to construct a non-uniform horizontal resolution over the computational domain. For example, the model resolution can be finer for coastal regions and coarser resolution for areas in the middle of the ocean that are not of interest. Therefore, irregular grids can be customised according to what the model is attempting to resolve. Furthermore, irregular grids can accuratley represent areas/regions of complex coastal morphology, bathymetery and physical complex nearshore systems. However, finite element modelling is not without its disadvantages, one being that these models struggle to resolve flow driven by geostrophy and advection, both of which are important characteristics of large-scale ocean flow. Issues also arise when modelling regions which vary in terms of viscosity and diffusivity coefficients. In addition, the issues of numerical diffusion in ocean models is compounded by spatially variable advection schemes when using irregular grids, making these issues harder to resolve. In addition, issues arise when trying to match regions of low resolution with regions of high resolution which imposes limitations on how ocean models resolve certain processes. However, irregular grids are common in coastal and estuarine models which need a high resolution of a specific, spatially complex area. Despite the limitations of using irregular grids for large scale ocean modelling, some predict that the challenges will be overcome and the use of regular grids for large scale modelling will become common place. For example, the development of an irregular, adaptive grid which dynamically changes the resolution as a function of flow in time.

In terms of vertical grids, there are also different 'types' that can be used, all of which have advantages and limitations. The vertical grid is sometimes reffered to as 'vertical coordinate system', and several aspects need to be taken into account when selecting a vertical grid system. For example one needs to take into account that most of the movement in the ocean is driven by movement at the surface, the ocean is strongly stratisfied, flow occurs along density gradients, and complex bathymetry over the computational domain. As mentioned previously, there are variety of vertical coordinate systems to suite a variety of applications with $Z$, $\theta$, and $\rho$ vertical coordinate systems regarded as the most popular. The absolute depth or $Z$-coordinate system is based on a series of depth levels and is easy to setup, is computationally efficient, and produces no pressure gradient errors. In addition, the resolution can be increased for a desired depth by simply reducing the spacing between the depth levels. However, when increasing the resolution along lateral boundaries (e.g. continental slopes) additional grid cells are needed. This leads to a similar problem as horizontal regular grids, where computational resources are wasted on areas which do not need to be resolved. The terrain following or $\theta$-coordinate system is based on *frictional* depth scaled from 0 to 1. For example, a 0.01$\theta$ level is 1% of the depth of the ocean, and 0.99$\theta$ level represents the 99% of the depth of the ocean. The advantage of this approach is the bathymetery is accurately represented, which allows high resolution near the seafloor, regardless of depth or proximity to land. However, a $\theta$-coordinate system does not resolve pressure gradients accurately as well as variable numerical advection schemes. The last popular approach is the isopycnal or density or $\rho$-coordinate system, which defines the grid based on density layers. Density layers are important drivers of flow in the ocean interior, which the $\rho$-coordinate system is able to characterise. The assumption made for this approach is ocean currents generally flow along surfaces of equal density for ocean interiors, meaning the flow is adiabatic. This approach works well for ocean interiors but performs poorly in shallow waters, where there is very little to no stratification in the water column. Therefore, in order for this approach to be applicable the fluid which flow is being resolved must be stratified. Since each of the approaches differ in terms of advantages and limitations, it is not uncommon to use a combination of coordinate systems known as hybrid coordinate systems. The hybrid coordinate system aims to optimise performance by combining the best suited coordinate system for a particular region. An example of a model that uses a hybrid coordinate system is the HYCOM model, which uses a combination of $Z$ coordinates for the surface layer and $\rho$ density in the ocean below. Hybrid coordinate systems allow vertical coordinates to evolve in both temporally and spatially as the depth changes. Although a dynamically optimised coordinate system improves the results of the model, it is done at a high computational cost.

Another important component of an ocean model is the resolution, in other words the size of the grid. Important to note is that the ocean exhibits 'multiple scale variability', which refers to the variability in time and length scales of the ocean. For example, vertical turbulent mixing occurs along a different time and length scale compared to the formation of eddies and fronts. The range of time and space scales can be from molecular (mm and seconds) to basin scale (10 000km and 1000 years). Furthermore, some processes are coupled across scales such that large scale processes are coupled with small scale processes, and are known as 'non-local interactions'. These non-local processes can have significant effects on both macro- and micro-scale components of the ocean. The resolution of a model is generally defined by the ability to permit or resolve eddies (see table below).

Resolution| Terminology| Definition
-----------|-------------|------------
$\ge$1$^\circ$ | Coarse | No eddies
$\sim$0.5$^\circ$ | Eddy-permitting| Some eddies
$\le$0.2$^\circ$ | Eddy-resolving| Eddies generate at a realistic strength and rate

Important to note is that the eddy 'resolving-model' does not resolve all eddies, or their assocaited aspects, but is a description of resolving eddies on what is currently known so far.

The parametrisation of processes in a model is another important component which must be accounted for to ensure computational resources are not wasted on processes which are negligible or too complex. In addition, the understanding of some processes include mesoscale eddy effects, dense overflows, submesoscale eddy effects, coastal processes, surface mixed layer process, friction, sub-grid scale mixing and ocean-ice interactions. Finally, the model needs to be validated to ensure it accurately represents a 'natural' ocean. Validation can be difficult as many observations are needed and the measurements themselves can be biased depending on the instruments used. Validation of ocean models has improved over the years due to initiatives such as the Argo Float Program. Other examples include satellites which are capable of measuring sea surface temperature, salinity and wind; while *in situ* instruments (buoys, ships, drifters, floats, temperature recorders, weather stations) provide direct measurements for validation purposes.

Ocean modelling is an important tool to help scientists understand how ocean processes are effected by climate change. However, ocean models are not perfect and challenges must be overcome to improve them, such as the effects of the internal wave field, model bias, dynamic ice sheets, mesoscale eddy fields, mixing and modelling drift. Furthermore, ocean modelling has provided a basis and information for the development of other tools such as hydrodynamic and trajectory modelling.
